On the effects of partying on next-day sports performance

### 2017-05-29 writing/draft.tex
Added explanation on previous research on alcohol and sports performance.

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
Anecdotally, it has been reported that professional sports teams party on 
the road, sometimes to the detriment of the game.\cite{hoch}
J.R. Smith is infamous for citing his improvement in performance due to a
lack of nightlife when moving to Cleveland from New York.\cite{price,ley}
A few interviews with former professional athletes revealed that location
matters more than day of the week, i.e. parties can happen any day of the week,
but they are more likely to occur in select cities.

We are interested to see if 
teams exhibit a decline in performance the day following a game in a city 
with active nightlife. We exploit data on bookmaker spreads, the expected score 
differential between two teams after conditioning on observable performance. 
We expect a team to meet the spread half the time, since this is how bookmakers 
minimize risk on their end. We construct a model which attempts to 
estimate the causal effect of  visiting a "party-city" on subsequent day 
performance as measured by the spread. Since next-day
opponent is uncorrelated with exposure to treatment, we have identification in our variable of 
interest.

\section{Previous Research}
In 2014, Ezekowitz tried to answer the question we're interested in.\cite{ezekowitz} 
However, the study only
looks at performance in party cities, as opposed to subsequent day performance (i.e.
a hangover effect). Moreover, their sample size was much smaller. Finally, there was no
methodology used behind the selection of party cities. Between these factors, no
conclusion was reached with statistical significance. The data do weakly suggest
that teams are less likely to win on the road in party cities.

In 2015, Chase also looked at a statistical ranking of the biggest partiers in
the NBA.\cite{chase}

Studies on effects of alcohol on sports performance in a controlled setting.
Same goes for sleep deprivation and possibly narcotics. 
Both of those well correlated with active night life.

\section{Materials and Methods}
One of the hardest parts of this analysis is that partying is a latent variable.
In order to Determine when players partied, one might consider to look at Twitter data
for direct mentions of partying, either by athletes or witnesses. However, due to
intense public scrutiny that athletes face, there is a strong reluctance to post
such mentions on social media. There are notable exceptions, such
as Oddel Beckham Jr.'s trip to Miami in 2017 season.\cite{bleler}
The reports are few and far enough between,
and could be biased based on whether a particular city has nightclubs with VIP rooms,
that picking up statistical evidence would be a fruitless.

>>>>

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
Anecdotally, it has been reported that professional sports teams party on 
the road, sometimes to the detriment of the game.\cite{hoch}
Within the NBA, J.R. Smith is infamous for citing his improvement in performance due to a
lack of nightlife when moving to Cleveland from New York.\cite{price,ley}
A few interviews with former professional athletes revealed that location
matters more than day of the week, i.e. parties can happen any day of the week,
but they are more likely to occur in select cities.

We are interested to see if 
teams exhibit a decline in performance the day following a game in a city 
with active nightlife. We exploit data on bookmaker spreads, the expected score 
differential between two teams after conditioning on observable performance. 
We expect a team to meet the spread half the time, since this is how bookmakers 
minimize risk on their end. We construct a model which attempts to 
estimate the causal effect of  visiting a ``party-city'' on subsequent day 
performance as measured by the spread. Since next-day
opponent is uncorrelated with exposure to treatment, we have identification in our variable of 
interest.

\section{Previous Research}
There has been an attempt to answer this question before.\cite{ezekowitz} 
However, the study only
looks at performance in party cities, as opposed to subsequent day performance (i.e. the study does not consider a delayed hangover effect). Moreover, their sample size was much smaller. 
Finally, there was no methodology used behind the selection of party cities. The author does conclude
that teams systematically play worse on Sunday afternoons when on the road,
regardless of cities.

Studies have also been carried out on statistical ranking of the players
most notorious for pursuing active nightlife in the NBA.\cite{chase} The study
does conclude that players perform worse on Sundays when compared with other days
as measured by the ratio of shots scored to shots taken.

Studies on effects of alcohol on sports performance in a controlled setting
show that alcohol impairs the nervous system, resulting in decrements in both
cognitive function and motor skill.\cite{shirreffs} In a study on 
the effects of alcohol on recovery of male athletes, it was found that
consumption is correlated with reduced production of testosterone
which ultimately leads to inhibited ability to recover and adapt to exercise.\cite{mjbarnes}
Same goes for sleep deprivation and possibly narcotics. 
These are all well correlated with active night life.

\section{Materials and Methods}
One of the hardest parts of this analysis is that partying is a latent variable.
In order to determine when players partied, one might consider to look at social media data
for direct mentions of partying, either by athletes or witnesses. However, due to
intense public scrutiny that athletes face, there is a strong reluctance to post
such mentions on social media. There are notable exceptions, such
as Oddel Beckham Jr.'s trip to Miami in 2017 season.\cite{bleler}
The reports are few and far enough between,
and could be biased based on whether a particular city has nightclubs with VIP rooms,
that picking up statistical evidence would be fruitless.
